Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-03-27 vertical-thermal-review.Rmd
moving yellow to Notes_S1

#Appendices
appendix_NEON_methods <- "S1"
appendix_tealeaves_methods <- "S2"
appendix_lit_review_methods <- "S3"

@@

Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).
Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`). 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022; @belandVerticalStructureHeterogeneity2021]. 
Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers. 
Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation, and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in forests with more open upper canopies, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
The implications of this light gradient are that leaf traits vary along the vertical gradient (Table 1), and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).

>>>>

#Appendices

Notes_S1<- "Notes S1"

appendix_NEON_methods <- "S1"
appendix_tealeaves_methods <- "S2"
appendix_lit_review_methods <- "S3"

@@

Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).
Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`, `r Notes_S1`). 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor (`r Notes_S1`).
Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers. 
Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation, and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in forests with more open upper canopies, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
The implications of this light gradient are that leaf traits vary along the vertical gradient (Table 1), and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).

### 2022-03-27 vertical-thermal-review_SI.Rmd
moving content to SI
moving yellow to Notes_S1

@@

#  `r appendix_NEON_methods_title`

>>>>

Notes_S1<- "Notes S1"

@@

# `r Notes_S1`

## Note S1: Vertical variation in forest structure across forests types

Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005]. 
In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].

The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022; @belandVerticalStructureHeterogeneity2021].

\newpage
#  `r appendix_NEON_methods_title`
